---
title: "Homework #02"
subtitle: "Statistical Methods in Data Science II & Lab"
author: ""
date: "**June 2022**"
output:
  html_document:
    keep_md: yes
    theme: united
  pdf_document:
    keep_tex: yes
    toc: no
header-includes: 
              - \usepackage[english]{babel}
              - \usepackage{amsmath}
              - \usepackage{enumerate}
              - \usepackage{setspace}
              - \usepackage{docmute}
              - \usepackage{fancyhdr}
              - \usepackage{graphicx}
              - \usepackage{rotating}
              - \usepackage{ucs}
              - \pagestyle{fancy}
              - \fancyhf{}
              - \rhead{Test \#01}
              - \cfoot{\thepage}
---


```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(expm)
library(invgamma)
library(R2jags)
library(LaplacesDemon)
library(corrplot)

set.seed(123)
knitr::opts_chunk$set(echo = TRUE)

# the default output hook
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x = unlist(stringr::str_split(x, '\n'))
    if (length(x) > n) {
      # truncate the output
      x = c(head(x, n), '....\n')
    }
    x = paste(x, collapse = '\n') # paste first n lines together
  }
  hook_output(x, options)
})
```

```{r, include=FALSE, include=FALSE, warning=FALSE}
opts_chunk$set(out.lines = 23)
```


```{r,echo=FALSE}
dugongs = read.delim2("./dugong-data.txt", header = TRUE, sep = "", dec = ",", stringsAsFactors = FALSE)
dugongs = dugongs[, -c(1)]
dugongs$Length = as.double(dugongs$Length)
dugongs$Age = as.double(dugongs$Age)
```



## Pinto Francesco - 1871045 

\vspace*{0cm}



\vspace*{1cm}





### 1.a Dataset exploration

Let's illustrate the characteristics of the statistical model for dealing with the *Dugong*'s data. Lengths ($Y_i$)  and  ages ($x_i$) of  27 dugongs ([sea cows](https://en.wikipedia.org/wiki/Dugong)) captured off the coast of Queensland have been recorded and the following (non linear)  regression model is considered in [Carlin and Gelfand (1991)](http://people.ee.duke.edu/~lcarin/Gelfand91.pdf):
\begin{eqnarray*}
Y_i &\sim& N(\mu_i, \tau^2) \\
\mu_i=f(x_i)&=& \alpha - \beta \gamma^{x_i}\\
\end{eqnarray*}
Model parameters are
$\alpha \in (1, \infty)$,
$\beta \in (1, \infty)$,
$\gamma \in (0,1)$,
$\tau^2 \in (0,\infty)$. 
Let us consider the following prior distributions:
\begin{eqnarray*}
\alpha &\sim&  N(0,\sigma^2_{\alpha})\\
\beta  &\sim&  N(0,\sigma^2_{\beta}) \\
\gamma &\sim&  Unif(0,1)\\
\tau^2 &\sim&  IG(a,b)) (Inverse Gamma)
\end{eqnarray*}


First of all let's analyze our sample by evaluating the sample mean and the sample variance:

```{r}
mu_i_hat = mean(dugongs$Length)
mu_i_hat

tau_2_hat = var(dugongs$Length)
tau_2_hat
```

And now we observe the frequency distribution of the observed outcome variables $Y$.

```{r, echo= F}
ggplot(dugongs, aes(x=Length)) + geom_histogram( fill=I("blue"), col=I("red"), alpha=I(.2), 
      breaks = seq(1.6, 2.8, 0.2)) + labs(title= 'Length of the Dugongs - Frequencies', x='Length', y='Count') 
```

We already know that the lengths have normal distribution, in fact we can observe a simil-normal distribution. Of course the "normality" is not evident because the sample is very small.


Instead, the age of the dugongs have the following distribution:




```{r, echo = F}
ggplot(dugongs, aes(x=Age)) + geom_histogram( fill=I("darkgreen"), col=I("orange"), alpha=I(.2), 
      breaks = seq(0, 35, 5)) + labs(title= 'Age of the Dugongs - Frequencies', x='Age', y='Count') 
```

We can see that the distribution is descent, because the longer is the life, the smaller is the probability to be still alive! 


And now the distribution of the length and age together:

```{r, echo = F}
ggplot(dugongs, aes(x=Age, y=Length)) + geom_point(col = 'magenta3') + labs( x='Age', y='Length')
```

At first glance it seems that the length and the age move together: we can study this phenomena by evaluating the correlation.

```{r}
cor = cor(dugongs$Length, dugongs$Age)
```

```{r, echo = F}
cat(paste("The correlation between x and Y is ", round(cor,4)*100, "%", sep =""))
```


The data don't follow a linear path, so we can suppose that the statistical model is defined by a polynomial regression.



### 1.b Derivation of the corresponding likelihood function

Given that our random variables $Y_i$ (the lengths) are independent, the general formula of the likelihood for our five parameters ($x_i, \alpha, \beta, \gamma, \tau^{2}$) is 

$$\mathcal{L}(x_i, \alpha, \beta, \gamma, \tau^{2}|y_i) = \prod_{i=1}^{n} f(y_i|x_i, \alpha, \beta, \gamma, \tau^{2})$$
That can be adapted to our distribution ($N(\mu_i, \tau^2)$) in the following way


$$\prod_{i=1}^{n} \frac{1}{\sqrt{2\pi\tau^{2}}} \cdot exp\Big\{-\frac{1}{2\tau^{2}} ( y_{i} - \mu_{i})^{2} \Big\}
= (2\pi\tau^{2})^{-\frac{n}{2}} \cdot exp\Big\{ -\frac{1}{2\tau^{2}} \sum_{i=1}^{n}\Big( y_{i} - (\alpha - \beta\gamma^{x_{i}})\Big)^{2} \Big\}$$



### 1.c Expression of the joint prior distribution of the parameters

Due to the fact that we are assuming independence between the parameters ($\alpha, \beta, \gamma, \tau^{2}$), the joint prior distribution will be the product of the respective priors.

$$\pi(\alpha, \beta, \gamma, \tau^{2}) =  f_{N(0,\sigma_\alpha^2)}(\alpha) \cdot f_{N(0,\sigma_\beta^2)}(\beta) \cdot f_{Unif(0, 1)}(\gamma) \cdot f_{IG(a, b)}(\tau^{2})$$
Before going on, let's quickly recap the respective distributions and try to exploit only the important information for the analysis (the information related to the hyperparameters):

$f_{N(0,\sigma_\alpha^2)}(\alpha) = \frac{1}{\sigma_\alpha\sqrt{2\pi}}\ exp\{\frac{-\alpha^2}{2\sigma_\alpha^2} \} \propto exp\{\frac{-\alpha^2}{2\sigma_\alpha^2} \}$ for $\alpha>1$

$f_{N(0,\sigma_\beta^2)}(\beta) = \frac{1}{\sigma_\beta\sqrt{2\pi}}\ exp\{\frac{-\beta^2}{2\sigma_\beta^2} \} \propto exp\{\frac{-\beta^2}{2\sigma_\beta^2} \}$ for $\beta>1$

$f_{Unif(0, 1)}(\gamma) = \frac{1}{1-0} = 1$ for $\gamma \in (0,1)$

$f_{IG(a, b)}(\tau^{2}) = \frac{b^a}{\Gamma(a)} \tau^{2^{(-a-1)}} e^{-\frac{b}{\tau^2}} \propto \tau^{2(-a-1)} exp\Big\{\frac{-b}{\tau^2}\Big\}$ for $\tau^2>0$

Then the final joint prior distribution can be written as 

$$\pi(\alpha, \beta, \gamma, \tau^{2}) = exp\Big\{\frac{-\alpha^2}{2\sigma_\alpha^2}\Big\} \cdot exp\Big\{\frac{-\beta^2}{2\sigma_\beta^2} \Big\} \cdot \tau^{2^{(-a-1)}} exp \Big\{ \frac{-b}{\tau^2} \Big\} \mathbb{I}_{\alpha}(1,\infty) \mathbb{I}_{\beta}(1,\infty) \mathbb{I}_{\tau^2}(1,\infty) \mathbb{I}_{\gamma}(0,1) = \\
 = \tau^{2^{(-a-1)}} exp \Big\{-\Big(\frac{\alpha^2}{2\sigma_\alpha^2} + \frac{\beta^2}{2\sigma_\beta^2} + \frac{b}{\tau^2} \Big) \Big\}\mathbb{I}_{\alpha}(1,\infty) \mathbb{I}_{\beta}(1,\infty) \mathbb{I}_{\tau^2}(0,\infty) \mathbb{I}_{\gamma}(0,1)$$



### 1.c' Suitable choice for the hyperparameters of $Y_i$

Now it's time to choose the best hyperparameters for $Y_i  \sim N(\mu_i, \tau^2)$

#### -Choice of $\mu_i$

In order to choose the best value of $\mu_i = \alpha - \beta \gamma^{x_i}$ we need to tune the best initial values for $\alpha$, $\beta$ and $\gamma$. These values will be modeled with the MCMC and with the jags functions.

The first step for determine an initial distribution for $\alpha$. Due to the fact that I start selecting $\alpha = 1.1$, I want a normal distribution that gives a good probability to this value:


```{r, echo = F}
curve(dnorm(x, 0, 5), col = 'orchid', xlim = c(-20,20), lwd = 3, ylim = c(0, 0.085), ylab = expression(pi(alpha)), xlab = expression(alpha))

for(i in seq(5,11,3))
  curve(dnorm(x, 0, i), col = 'grey', add = T)
abline(v = 1.1, lty = 2, col = 'orange')

legend(9, 0.08, legend=c("Best prediction", "Other predictions"), col=c("orchid", "grey"), lty=1, cex=0.8)
grid()
```

My initial choice for $\sigma_\alpha^2$ is $5$, so $\alpha \sim N(0,5)$

Then I evaluate $\sigma_\beta^2$ such as it gives a good plausibility to my initial choice $\beta = 1.1$. Also in this case I choose the same distribution as before, $\beta \sim N(0,5)$

```{r, echo = F}
curve(dnorm(x, 0, 5), col = 'darkblue', xlim = c(-20,20), lwd = 3, ylim = c(0, 0.085), ylab = expression(pi(beta)), xlab = expression(beta))

for(i in seq(5,11,3))
  curve(dnorm(x, 0, i), col = 'grey', add = T)
abline(v = 1.1, lty = 2, col = 'orange')

legend(9, 0.08, legend=c("Best prediction", "Other predictions"), col=c("darkblue", "grey"), lty=1, cex=0.8)
grid()
```



Regarding the parameter $\gamma$ I don't need to find its hyperparameters, because its distribution is already known: $\gamma \sim Unif(0,1)$

#### -Choice of $\tau^2$

To choose the best $\tau^2$ for our model, I decided to study the uncertainty by looking at the variances for different values of $a \in (2,10]$ and $b \in [1,10]$. I did a matrix that contains the variances for every couple of $a$ and $b$ and then extracted the best parameters (those with smallest variance). 

[**Reminder**] the variance of an Inverse Gamma is  

$$\sigma_{IG}^2= \frac{b^2}{(a-1)^2(a-2)}$$

```{r}
#values for a
a = seq(2.001,10,0.1)

#values for b
b = seq(1,10,0.1)

#creation of the matrix
x = matrix(data = NA, nrow = length(a), ncol = length(b))

#function of the variance
var = function(a,b) b^2/((a-1)^2 * (a-2))

#matrix with the variances for all the possible values a and b
for(i in 1:length(a)){
  for(j in 1:length(b)){
    x[i,j] = var(a[i], b[j])
  }
}

#hyperparameters for the smallest variance
#which command finds the required values for row and column (in  this case the min)
a_min = a[which(x==min(x), arr.ind=T)[1]] 
b_min = b[which(x==min(x), arr.ind=T)[2]]
```

```{r, echo = F}
cat(paste("The best hyperparameters for ", 'τ^2 ',
          'are a=', a_min, " and b=", b_min, sep =""))
```


```{r, echo = F}
curve(dinvgamma(x, a_min, b_min), xlim = c(0,0.3), ylim = c(0,13), col = 'orchid',
      xlab = 'x', ylab = expression(pi(tau^2)), main = expression(paste("PDF of ", tau^2, " as InvGamma(9.901, 1)")), lwd = 3)

legend(0.21, 12, legend=c("Best prediction", "Other predictions"), col=c("orchid", "grey"), lty=1, cex=0.8)

grid()

for(a in c(3,6, 10)){
  for(b in c(1.5, 2)){
    curve(dinvgamma(x, a, b), col = 'grey', add = T)
  }
}
```





### 1.d Derivation of the functional form  (up to proportionality constants) of all *full-conditionals*

The first full conditionals regard the distribution of $\mu_i$, so I start analyzing the parameters $\alpha$, $\beta$ and $\gamma$:

**-1 full conditional: $\alpha$**

$$
\pi(\alpha|\beta,\gamma,\mu_i,\tau^2,y_0) \propto \pi(\mu_i|\alpha, \beta, \gamma)\pi(\alpha) 
\propto\\ \propto (2\pi\tau^{2})^{-\frac{n}{2}}  exp\Big\{ -\frac{1}{2\tau^{2}} \sum_{i=1}^{n}\Big( y_{i} - (\alpha - \beta\gamma^{x_{i}})\Big)^{2} \Big\} exp\{\frac{-\alpha^2}{2\sigma_\alpha^2} \} \\
 \propto  \tau^{2 (-\frac{n}{2})} exp\Big\{ -\frac{1}{2\tau^{2}} \Big [\sum_{i=1}^{n}\Big( y_{i} - (\alpha - \beta\gamma^{x_{i}})\Big)^{2}  - \frac{\alpha^2}{\sigma_{\alpha}^2}\Big ] \Big\}
$$

for $\alpha>1$

**-2 full conditional: $\beta$**

$$
\pi(\beta|\alpha,\gamma,\mu_i,\tau^2,y_0) \propto \pi(\mu_i|\alpha, \beta, \gamma)\pi(\beta) 
 \propto\\ \propto (2\pi\tau^{2})^{-\frac{n}{2}}  exp\Big\{ -\frac{1}{2\tau^{2}} \sum_{i=1}^{n}\Big( y_{i} - (\alpha - \beta\gamma^{x_{i}})\Big)^{2} \Big\} exp\{\frac{-\beta^2}{2\sigma_\beta^2} \} \propto \\
 \propto  \tau^{2 (-\frac{n}{2})} exp\Big\{ -\frac{1}{2\tau^{2}} \Big [\sum_{i=1}^{n}\Big( y_{i} - (\alpha - \beta\gamma^{x_{i}})\Big)^{2}  - \frac{\beta^2}{\sigma_{\beta}^2}\Big ] \Big\}
 \\
$$

for $\beta>1$

**-3 full conditional: $\gamma$**

$$
\pi(\gamma|\alpha,\beta,\mu_i,\tau^2,y_0) \propto \pi(\mu_i|\alpha, \beta, \gamma)\pi(\gamma)
\\ 
\propto exp\Big\{ -\frac{1}{2\tau^{2}} \sum_{i=1}^{n}\Big( y_{i} - (\alpha - \beta\gamma^{x_{i}})\Big)^{2} \Big\} \mathbb{I}_{\gamma}(0,1)
$$


And then I evaluate the full conditional for $\tau^2$:


$$
\pi(\tau^2|\alpha,\beta,\mu_i,y_0) \propto \pi(y_0|\alpha, \beta, \gamma, \tau^2)\pi(\tau^2)\propto \pi(y_0|\mu_i, \tau^2) \pi(\tau^2) \propto\\
\propto (2\pi\tau^{2})^{-\frac{n}{2}}  exp\Big\{ -\frac{1}{2\tau^{2}} \sum_{i=1}^{n}\Big( y_{i} - (\alpha - \beta\gamma^{x_{i}})\Big)^{2} \Big\}\tau^{2(-a-1)} exp\Big\{\frac{-b}{\tau^2}\Big\}\\
\propto\tau^{2(-\frac{n}{2}-a-1)}exp\Big\{ -\frac{1}{2\tau^{2}} \sum_{i=1}^{n}\Big( y_{i} - (\alpha - \beta\gamma^{x_{i}})\Big)^{2} -\frac{b}{\tau^2} \Big\} = \\
\tau^{2(-\frac{n}{2}-a-1)}exp\Big\{ -\frac{1}{2\tau^{2}} \Big [\sum_{i=1}^{n}\Big( y_{i} - (\alpha - \beta\gamma^{x_{i}})\Big)^{2} -2b \Big ] \Big\}
$$

for $\tau^2>0$



### 1.e The respective distributions

I can recognize the following distributions from the formulas above:

$$
\alpha \sim N\Bigg (\mu = \frac{\sum_{i=1}^n (y_i + \beta\gamma^{x_i})}{\tau^2\Big(\frac{1}{\sigma_{\alpha}^2}+\frac{n}{\tau^2}\Big)}, \sigma^2 = \frac{1}{\frac{1}{\sigma_\alpha^2}+\frac{n}{\tau^2}}  \Bigg )
$$
$$
\beta \sim N\Bigg (\mu = \frac{\sum_{i=1}^n (y_i\gamma^{x_i} + \alpha\gamma^{x_i})}{\tau^2\Big(\frac{1}{\sigma_{\alpha}^2}+\frac{\sum_{i=1}^n\gamma^{2x_i}}{\tau^2}\Big)}, \sigma^2 = \frac{1}{\frac{1}{\sigma_\alpha^2}+\frac{\sum_{i=1}^n\gamma^{2x_i}}{\tau^2}}  \Bigg )
$$

$$
\tau^2 \sim IG \Big (a = a+\frac{n}{2}+1, b = \frac{1}{2}\sum_{i=1}^n(y_i - \alpha + \beta\gamma^{x_i})^2 + b \Big )
$$

### 1.f Markov Chain simulation with Metropolis-within-Gibbs algorithm

Doing a MC simulation manually (from scratch) with Metropolis-within-Gibbs algorithm would be so long and time-consuming. Luckily there is the jags package that does all the passage automatically for us! So I'm gonna do the simulation as follows.

```{r, echo = F}
mydata <- list(   x = c( 1.0,  1.5,  1.5,  1.5, 2.5,   4.0,  5.0,  5.0,  7.0,
                      8.0,  8.5,  9.0,  9.5, 9.5,  10.0, 12.0, 12.0, 13.0,
                      13.0, 14.5, 15.5, 15.5, 16.5, 17.0, 22.5, 29.0, 31.5),
               Y = c(1.80, 1.85, 1.87, 1.77, 2.02, 2.27, 2.15, 2.26, 2.47,
                     2.19, 2.26, 2.40, 2.39, 2.41, 2.50, 2.32, 2.32, 2.43,
                     2.47, 2.56, 2.65, 2.47, 2.64, 2.56, 2.70, 2.72, 2.57), N = 27)
```


```{r}
t = 10000 #steps
parameters = c('alpha', 'beta', 'gamma', 'sigma') 
#due to the fact that invgamma is not supported in the model file for jags, I'm gonna study the parameter sigma=1/sqrt(tau), where tau has gamma distribution (like the real tau^2). 
#I'm gonna consider 'sigma^2' as tau^2 where sigma is the inverse of a gamma

inits = list(alpha = 1.1, beta = 1.1, tau_2 = 0.5, gamma = 1) #initial parameters mentioned above
inits = list(inits) #converting them to list in order to become readable for jags

djags = jags(data = mydata, inits = inits, parameters.to.save = parameters, model.file = '2022-jags-model-dugong.txt', n.chains = 1, n.thin = 1, n.burnin = 10, n.iter = t)
```


### 1.g  Trace-plots of the simulations of each parameter

I need the trace-plot for my parameters to make sure that my distribution is well calibrated.

In fact, we can observe that the distribution of the Markov Chain for every parameter is balanced around the mean, that is represented with the dashed line:

```{r, echo = F}
plot(djags$BUGSoutput$sims.array[, 1, "alpha"],type = 'l', col = 'darkred', main = expression(paste("Trace-plot of the Markov Chain for ", alpha)), xlab ='t', ylab = expression(alpha))
abline(h = mean(djags$BUGSoutput$sims.array[, 1, "alpha"]), lty = 2, lwd = 3, col = 'orange')
grid()
```


```{r, echo = F}
plot(djags$BUGSoutput$sims.array[,1,"beta"],type = 'l', col = 'orange', main = expression(paste("Trace-plot of the Markov Chain for ", beta)), xlab ='t', ylab = expression(beta))#, ylim = c(0.5,1.6))
abline(h = mean(djags$BUGSoutput$sims.array[, 1, "beta"]), lty = 2, lwd = 3, col = 'orchid')
grid()
```

```{r, echo = F}
plot(djags$BUGSoutput$sims.array[,1,"gamma"],type = 'l', col = 'darkgreen', main = expression(paste("Trace-plot of the Markov Chain for ", gamma)), xlab ='t', ylab = expression(gamma))#, ylim = c(0.6,1.1))
abline(h = mean(djags$BUGSoutput$sims.array[, 1, "gamma"]), lty = 2, lwd = 3, col = 'orange')
grid()
```


```{r, echo = F}
plot(djags$BUGSoutput$sims.array[,1,"sigma"],type = 'l', col = 'orchid', main = expression(paste("Trace-plot of the Markov Chain for ", sigma)), xlab ='t', ylab = expression(sigma))#, ylim = c(0.05,0.23))
abline(h = mean(djags$BUGSoutput$sims.array[, 1, "sigma"]), lty = 2, lwd = 3, col = 'orange')
grid()
```


### 1.h  Graphical evaluation of the empirical averages $\hat{I}_t$  with growing $t=1,...,T$

If $I<\infty$ and $\theta_1,...\theta_t$ iid, for the Strong Law of Large Numbers (SLLN), the empirical average is a consistent (sequence of) estimator(s) of $I$, such as 
$$
\hat{I}_{t} = \frac{1}{T} \sum_{i=1}^{T} h(\theta_i)\xrightarrow{\text{a.s.}} I
$$
So now I implement the estimator in the following way and apply it to every parameter.

```{r}
n = length(djags$BUGSoutput$sims.array[,,1])
sum = 0
I_t = rep(NA, n)

#this is the singular case of alpha ([i,,1] represent the alpha index)
for(i in 1:n){
  sum = sum + djags$BUGSoutput$sims.array[i,,1] 
  I_t[i] = sum/i
}
```



```{r, echo = F}
plot(I_t, type = 'l', col = 'orange', xlab = 't', ylab = expression(alpha), main = expression(paste('Empirical Average over t iterations for ', alpha) ))
grid()

sum = 0
I_t = rep(NA, n)

for(i in 1:n){
  sum = sum + djags$BUGSoutput$sims.array[i,,2]
  I_t[i] = sum/i
}
plot(I_t, type = 'l', col = 'orchid', xlab = 't', ylab = expression(beta), main = expression(paste('Empirical Average over t iterations for ', beta) ))
grid()

sum = 0
I_t = rep(NA, n)

for(i in 1:n){
  sum = sum + djags$BUGSoutput$sims.array[i,,4]
  I_t[i] = sum/i
}
plot(I_t, type = 'l', col = 'green', xlab = 't', ylab = expression(gamma), main = expression(paste('Empirical Average over t iterations for ', gamma) ))
grid()

sum = 0
I_t = rep(NA, n)

for(i in 1:n){
  sum = sum + djags$BUGSoutput$sims.array[i,,5]
  I_t[i] = sum/i
}
plot(I_t, type = 'l', col = 'red', xlab = 't', ylab = expression(sigma), main = expression(paste('Empirical Average over t iterations for ', sigma) ))
grid()
```


### 1.i Provide estimates for each parameter together with the approximation error and explain how you have evaluated such error

I can simply extract the estimates for each parameter by looking into the table generated by the jags function as follows:

The estimate average of $\alpha$ is 
```{r}
as.double(djags$BUGSoutput$mean['alpha'])
```

for $\beta$ is
```{r}
as.double(djags$BUGSoutput$mean['beta'])
```


for $\gamma$ is
```{r}
as.double(djags$BUGSoutput$mean['gamma'])
```

and for $\sigma$ is 
```{r}
as.double(djags$BUGSoutput$mean['sigma'])
```

To evaluate the approximation error I use the **MCSE** estimate: the MCSE (Monte Carlo Standard Error) is an estimate of the inaccuracy of Monte Carlo samples, usually regarding the expectation of posterior samples from  Monte Carlo Markov Chain algorithms.

I estimate the MCSE with the $MCSE$ function from LaplacesDemon package

```{r}
#AE = Approximation Rrror
AE_alpha = MCSE(djags$BUGSoutput$sims.array[,1,"alpha"])
AE_beta = MCSE(djags$BUGSoutput$sims.array[,1,"beta"])
AE_gamma = MCSE(djags$BUGSoutput$sims.array[,1,"gamma"])
AE_sigma = MCSE(djags$BUGSoutput$sims.array[,1,"sigma"])
```


```{r, echo = F}

cat('The approximation error for alpha is ', AE_alpha, '\n',
    'The approximation error for beta is ', AE_beta, '\n',
    'The approximation error for gamma is ', AE_gamma, '\n',
    'The approximation error for sigma is ', AE_sigma)
```

### 1.l  Which parameter has the largest posterior uncertainty? How did you measure it?

To measure the posterior uncertainty, a good indicator could be the standard deviation. 

The uncertainty is sometimes contained into the variability of the data that distorts a good prediction for the parameters.

```{r}
sd_alpha = djags$BUGSoutput$sd$alpha
sd_beta = djags$BUGSoutput$sd$beta
sd_gamma = djags$BUGSoutput$sd$gamma
sd_sigma = djags$BUGSoutput$sd$sigma
```

```{r, echo = F}
cat('The posterior uncertainty alpha is ', sd_alpha, '\n',
    'The posterior uncertainty for beta is ', sd_beta, '\n',
    'The posterior uncertainty for gamma is ', sd_gamma, '\n',
    'The posterior uncertainty for sigma is ', sd_sigma)
```

At a first glance, it seems to be $\alpha$ the parameter with highest posterior uncertainty!

### 1.m  Parameters with the largest correlation 

To discover the parameters with highest correlation, I simply evaluate a correlation plot between them:

We can clearly see that all the parameters have correlation (positive or negative), but the highest is the one between $\alpha$ and $\gamma$ and it's a positive correlation.

```{r, echo = F}
corrplot(cor(djags$BUGSoutput$sims.matrix)[-3,-3], method = "color", addCoef.col="black", col = COL2('RdYlBu'))
```



### 1.n  Posterior predictive  distribution of the length of a dugong with age of 20 years.

Before the data are considered, the distribution of the unknown $Y$ is 

$$m(y) = \int_{\Theta}f(y|\theta)d\theta$$


It's the marginal distribution of $y$ and it's called *prior predictive distribution*. With this distribution we can only try to do predictions with the information given by the single parameter.

After observing the data $Y=(Y_1,...,Y_n) = y = (y_1,...,y_n)$, the **posterior predictive distribution** is given by 
$$m(y_{new}|y_1,...,y_n) = \int_\Theta f(y_{new}|\theta)\pi(\theta|\textbf{y})$$
It is the distribution of possible unobserved values conditional on the observed values.

In the MCMC case this formula can be avoided and, also in this case, I can rely on the jags function

```{r, results='hide', warning = F}
pred_20 <- jags(data=mydata, inits=inits, parameters.to.save = c('Ypred20'), model.file = 'Model_with_predictions.txt', n.chains=1,n.iter=10000,n.thin = 1,n.burnin = 1)
#Ypred20 stands for the prediction of Y for x = 20
```

```{r, echo = F}
cat('The posterior predictive distribution for a dugong with age of 20 years suggests us that his length will be', round(as.double((pred_20$BUGSoutput$mean$Ypred20)), 2), 'meters.')
```


### 1.o  Prediction of a different dugong with age 30 

Using the same method as before,

```{r, results='hide', warning = F}
pred_30 <- jags(data=mydata, inits=inits, parameters.to.save = c('Ypred30'), model.file = 'Model_with_predictions.txt', n.chains=1,n.iter=10000,n.thin = 1,n.burnin = 1)
#Ypred30 stands for the prediction of Y for x = 30
```

```{r, echo = F}
cat('The posterior predictive distribution for a dugong with age of 30 years suggests us that his length will be', round(as.double((pred_30$BUGSoutput$mean$Ypred30)), 2), 'meters.')
```

### 1.p The less precise prediction

What's the less reliable between the last two predictions? We have two ways to evaluate this factor: the standard deviation and the width of the Equal-Tail interval.

**Standard Deviaton:** I consider less reliable the prediction with higher standard deviation:

```{r}
sd_20 = as.double(pred_20$BUGSoutput$sd[1])
sd_30 = as.double(pred_30$BUGSoutput$sd[1])
```

```{r, echo = F}
if(sd_20>sd_30)
  cat('The standard deviation of the first prediction is higher, (',sd_20 ,'), so the second prediction is better (with sd=',sd_30 ,')')
if(sd_30>sd_20)
  cat('The standard deviation of the second prediction is higher, (',sd_30 ,'), so the first prediction is better (with sd=',sd_20 ,')')
```


**ET-Interval:** The longest is the interval, the less reliable is the prediction, because the data are more distributed and so the probabilities are not much high on a specific point. In this case my interval is at level $\alpha = 0.05$:

```{r}
length_20 = pred_20$BUGSoutput$summary[1,'97.5%'] - pred_20$BUGSoutput$summary[1,'2.5%']
length_30 = pred_30$BUGSoutput$summary[1,'97.5%'] - pred_30$BUGSoutput$summary[1,'2.5%']
```

```{r, echo = F}
if(length_20>length_30)
  cat('The ET-interval range of the first prediction is higher, (',length_20 ,'), so the second prediction is better (with ET range of',length_30 ,')')
if(length_30>length_20)
  cat('The ET-interval range of the second prediction is higher, (',length_30 ,'), so the first prediction is better (with ET range of',length_20 ,')')
```

In both case we prefer the first prediction (on 20 y.o. dugongs), so we consider it as the best one!

\newpage

# Part 2 - Markov Chain

Let us consider a Markov Chain 
$(X_t)_{t \geq 0}$
defined on the state space ${\cal S}=\{1,2,3\}$
with the following transition matrix


\begin{pmatrix}
0 & 1/2 & 1/2\\
5/8 & 1/8 & 1/4\\
2/3 & 1/3 & 0
\end{pmatrix}


### 2.a Markov Chain after 1000 steps

My goal is to find the distribution of the Markov Chain starting at time $t=0$ in the state  $X_0=1$ for $t=1000$ consecutive times.


**Step 1:** implementation of the transition matrix in R
```{r}
#Implementation of the transition matrix:
P = matrix(data = c(0, 5/8, 2/3, 1/2, 1/8, 1/3, 1/2, 1/4, 0), 
             nrow=3, ncol=3)

#Rounded matrix (some values are periodical numbers, so we round them in the visualization of the matrix):
print(round(P,2)) 
```

**Step 2:** evaluation of the MC after 1000 steps.

```{r}
S=c(1, 2, 3) #states

x0 = 1 
n = 1000
chain = rep(NA, n+1) #implementation of the chain matrixx

chain[1] = x0 #first step          

t = 0
for(t in 1:n){
  chain[t+1] = sample(x = S, size = 1, prob = P[chain[t],])
}
```

### 2.b Empirical Relative Frequency

The empirical relative frequencies of the states in my simulation are the following:

```{r}
prop.table(table(chain))
```


### 2.c Simulation pt. 2

I repeat the simulation for $500$ times and record only the final state at time $t=1000$ for each of the  500 simulated chains, computing the relative frequency of the 500 final states.


```{r}
val = rep(0,500)

for(i in 1:500){
  t =  0
  for(t in 1:n){
    chain[t+1] = sample(x=S, size=1, prob=P[chain[t],])
  }
  val[i] = chain[1000]
}

prop.table(table(val))
```


What distribution are you approximating in this way?  
Try to formalize the difference between this point and the previous
point. 


### 2.d Stationary distribution

Given $\pi$, probability vector over the state space $S$ (such as $\sum_{i\in S}\pi_i=1$ and $\pi_i \geq 0$ $\forall i \in S$) we consider $\pi$ as the Stationary Distribution of the Markov Chain if 

$$\pi=\pi P$$
Where $P$ is the transition matrix

so, in other words, $\pi_j=\sum_{i\in S}\pi_i p_{ij}$ for every $j \in S$

Given that, we can find the stationary distribution in two equivalent ways: by solving the system below, or by multiplying the transition matrix $P$ itself infinite times. Let's do both:

**System resolution**

\begin{cases}
\pi_1=\frac{5}{8}\pi_2+\frac{2}{3}\pi_3\\
\pi_2=\frac{1}{2}\pi_1+\frac{1}{8}\pi_2+\frac{1}{3}\pi_3\\
\pi_3=\frac{1}{2}\pi_1+\frac{1}{4}\pi_2\\
\pi_1+\pi_2+\pi_3=1
\end{cases}

The system returns the following (rounded) solutions: $\pi_1=0.392$, $\pi_2=0.330$, $\pi_3 = 0.278$, so $\pi = (0.392, 0.330, 0.278)$.

**Matrix-multiplication resolution**

$$\pi = \lim_{n\rightarrow \infty}P^n$$

I decide to evaluate $\pi$ by elevating the $P$ matrix to an high integer, because the result remains the same (for example if $n$ is $100$, $1000$ or $10000$ the stationary distribution remains the same)


```{r}
P_1000 = P%^%1000 #this function is taken from expn package
P_1000[1,]
```

The result is the same. I'm satisfied of my calculations.

### 2.e Well-approximation of the ERF

Is the ERF well approximated by the simulated empirical relative frequencies computed in 1.b and 1.c?
  
in the 1.b I computed:
```{r, echo = F}
prop.table(table(chain))
```
  
In the point 1.c I computed  

```{r, echo = F}
prop.table(table(val))
```
  
And then the real ERF is 

```{r, echo = F}
P_1000[1,]
```


We can conlude that they are very similar!


### 2.f What if...?

What happens if we start at $t=0$ from state  $X_0=2$ instead of $X_0=1$?


```{r}
x0 = 2 
chain = rep(NA, n+1) 

chain[1] = x0 #first step          
t = 0

for(t in 1:n){
  chain[t+1] = sample(x = S, size = 1, prob = P[chain[t],])
}
```

```{r, echo = F}
prop.table(table(chain))
```


Of course the distribution has no relevant changes because, on the long period, the MC will have the same behaviour and so the same frequencies.


\vspace{6cm}




